---
title: "COVID-19 in Newfoundland and Labrador"
author: "Dr. Amy Hurford"
output:
  html_document: default
  word_document: default
  pdf_document: default
fontsize: 11pt
---
[Memorial University of Newfoundland and Labrador,](https://www.mun.ca/math/)
[Canadian Network for Modelling Infectious Disease,](https://canmod.net/)
[Mathematics for Public Health,](http://www.fields.utoronto.ca/activities/public-health)
[One Health Modelling Network for Emerging Infections](https://www.yorku.ca/science/cdm/2021/04/09/ccdm-network-to-model-emerging-infectious-diseases-receives-2-5-million-in-federal-funding/)

```{r setup, include=FALSE}
require(zoo)
require(ggplot2)
require(patchwork)
gg_color_hue <- function(n){
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}
n = 6
cols = gg_color_hue(n)
#data=read.csv("~/Desktop/RHA_DailyData_Original_Public.csv")
data <- read.csv('https://raw.githubusercontent.com/ahurford/covid-nl/master/NL-Hub/RHA_DailyData_Original_Public.csv', fill=TRUE)
data1<-data
data2 = data.frame(date=as.Date(data$date_updated), cases = data$new_cases, tests = data$total_people_tested, RHA = data$regional_health_authority)
data2 = data2[data2$date>"2021-12-10",]
eastern = data2[data2$RHA=="eastern_rha",]
western = data2[data2$RHA=="western_rha",]
central = data2[data2$RHA=="central_rha",]
labrador = data2[data2$RHA=="labradorgrenfell_rha",]

row.names(eastern) <- NULL
row.names(central)<-NULL
row.names(western)<-NULL
row.names(labrador)<-NULL

# correct flaws in data
eastern$tests[31] = eastern$tests[31]+eastern$tests[30]
# repeated days or non-sensical test numbers
eastern = eastern[-c(44,28,35),]
western = western[-43,]
central = central[-c(44,28),]
# Number of tests not reported on entry 38
labrador = labrador[-c(44,28,38),]

smoothdata = function(){
T = length(values[,1])
difference <- c(1,as.double(difftime(values$date[2:T],values$date[1:(T-1)],units='days')))
# Cases per day since last reporting
values$tests = diff(c(0,values$tests))/difference
values$cases = values$cases/difference
values = cbind(values,difference=difference,positivity=values$cases/values$tests)
dates = seq(as.Date("2021-12-14"), values$date[T], by="days")

cases = NULL
tests = NULL
positivity = NULL

ii = NULL
for(i in seq(1,(T-1))){
  cases = c(cases,rep(values$cases[(i+1)],difference[(i+1)]))
  tests = c(tests,rep(values$tests[(i+1)],difference[(i+1)]))
  positivity = c(positivity,rep(values$positivity[(i+1)],difference[(i+1)]))
}

cases = tail(cases,-1)
tests = tail(tests,-1)
positivity = tail(positivity,-1)

values = data.frame(dates=tail(dates,-1), cases=cases, tests=tests, positivity = positivity)

# 14 day rolling mean:
values.roll = data.frame(cases = rollmean(cases,14), tests = rollmean(tests,14), positivity = rollmean(positivity,14))
n=0.5
undetected = popn.size*((values.roll$cases/popn.size)^n*(values.roll$positivity)^(1-n))


cases.raw = values$cases
test.raw = values$tests

data = data.frame(dates2 = tail(dates,-1), tests.roll = c(rep(NA,13),values.roll$tests), detected.roll = c(rep(NA,13),values.roll$cases), all = c(rep(NA,7),undetected+values.roll$cases, rep(NA,6)), cases.raw = cases.raw, test.raw = test.raw, positivity.roll = c(rep(NA,13),values.roll$positivity))
return(data)
}

values <- eastern
popn.size <- 10000*data1[data1$regional_health_authority=="eastern_rha",]$Population[1]
data = smoothdata()
eastern = data

gE=ggplot() +
  #geom_point(data=data, aes(x=as.Date(dates2), y=100*cases.raw/max(cases.raw)), col = cols[1], size=1)+
#geom_point(data=data, aes(x=as.Date(dates2), y=100*(cases.raw/test.raw)/max(cases.raw/test.raw)), col = "grey", size=1)+
  geom_point(data=data[!is.na(data$detected.roll),], aes(x=as.Date(dates2), y=100*detected.roll/max(detected.roll)), col = cols[1], size=1)+
  geom_point(data=data[!is.na(data$positivity.roll),],aes(x=as.Date(dates2), y=100*positivity.roll/max(positivity.roll)),size=1,col = "grey")+
  geom_line(data=data[!is.na(data$all),],aes(x=as.Date(dates2), y=100*all/max(all)),lwd=2, col = cols[1])+
  xlab("")+
  ylab("% of maximum")+
  ggtitle("Eastern Health")+
  theme_classic()+
  scale_x_date(date_breaks = "7 day", date_labels = "%d %b")+
  theme(axis.text.x = element_text(angle = 90),legend.position = "none")
##

values <- central
popn.size <- 10000*data1[data1$regional_health_authority=="central_rha",]$Population[1]
data = smoothdata()
central = data

gC=ggplot() +
  #geom_point(data=data, aes(x=as.Date(dates2), y=100*cases.raw/max(cases.raw)), col = cols[1], size=1)+
#geom_point(data=data, aes(x=as.Date(dates2), y=100*(cases.raw/test.raw)/max(cases.raw/test.raw)), col = "grey", size=1)+
  geom_point(data=data[!is.na(data$detected.roll),], aes(x=as.Date(dates2), y=100*detected.roll/max(detected.roll)), col = cols[2], size=1)+
  geom_point(data=data[!is.na(data$positivity.roll),],aes(x=as.Date(dates2), y=100*positivity.roll/max(positivity.roll)),size=1,col = "grey")+
  geom_line(data=data[!is.na(data$all),],aes(x=as.Date(dates2), y=100*all/max(all)),lwd=2, col = cols[2])+
  xlab("")+
  ylab("% of maximum")+
  ggtitle("Central Health")+
  theme_classic()+
  scale_x_date(date_breaks = "7 day", date_labels = "%d %b")+
  theme(axis.text.x = element_text(angle = 90),legend.position = "none")
##

values <- western
popn.size <- 10000*data1[data1$regional_health_authority=="western_rha",]$Population[1]
data = smoothdata()
western = data

gW=ggplot() +
  #geom_point(data=data, aes(x=as.Date(dates2), y=100*cases.raw/max(cases.raw)), col = cols[1], size=1)+
#geom_point(data=data, aes(x=as.Date(dates2), y=100*(cases.raw/test.raw)/max(cases.raw/test.raw)), col = "grey", size=1)+
  geom_point(data=data[!is.na(data$detected.roll),], aes(x=as.Date(dates2), y=100*detected.roll/max(detected.roll)), col = cols[3], size=1)+
  geom_point(data=data[!is.na(data$positivity.roll),],aes(x=as.Date(dates2), y=100*positivity.roll/max(positivity.roll)),size=1,col = "grey")+
  geom_line(data=data[!is.na(data$all),],aes(x=as.Date(dates2), y=100*all/max(all)),lwd=2, col = cols[3])+
  xlab("")+
  ylab("% of maximum")+
  ggtitle("Western Health")+
  theme_classic()+
  scale_x_date(date_breaks = "7 day", date_labels = "%d %b")+
  theme(axis.text.x = element_text(angle = 90),legend.position = "none")
##

values <- labrador
popn.size <- 10000*data1[data1$regional_health_authority=="labradorgrenfell_rha",]$Population[1]
data = smoothdata()
labrador = data

gL=ggplot() +
  #geom_point(data=data, aes(x=as.Date(dates2), y=100*cases.raw/max(cases.raw)), col = cols[1], size=1)+
#geom_point(data=data, aes(x=as.Date(dates2), y=100*(cases.raw/test.raw)/max(cases.raw/test.raw)), col = "grey", size=1)+
  geom_point(data=data[!is.na(data$detected.roll),], aes(x=as.Date(dates2), y=100*detected.roll/max(detected.roll)), col = cols[4], size=1)+
  geom_point(data=data[!is.na(data$positivity.roll),],aes(x=as.Date(dates2), y=100*positivity.roll/max(positivity.roll)),size=1,col = "grey")+
  geom_line(data=data[!is.na(data$all),],aes(x=as.Date(dates2), y=100*all/max(all)),lwd=2, col = cols[4])+
  xlab("")+
  ylab("% of maximum")+
  ggtitle("Labrador-Grenfell Health")+
  theme_classic()+
  scale_x_date(date_breaks = "7 day", date_labels = "%d %b")+
  theme(axis.text.x = element_text(angle = 90),legend.position = "none")
##

cases.raw = eastern$cases.raw + western$cases.raw + central$cases.raw + labrador$cases.raw
test.raw = eastern$test.raw + western$test.raw + central$test.raw + labrador$test.raw
all = eastern$all + western$all + central$all + labrador$all
positivity.roll = c(rep(NA,13),rollmean(cases.raw/test.raw, 14))
detected.roll = c(rep(NA,13),rollmean(cases.raw, 14))
data = data.frame(dates2 = labrador$dates2, cases.raw=cases.raw, detected.roll = detected.roll, all = all, positivity.roll = positivity.roll)


gNL=ggplot() +
  #geom_point(data=data, aes(x=as.Date(dates2), y=100*cases.raw/max(cases.raw)), col = cols[1], size=1)+
#geom_point(data=data, aes(x=as.Date(dates2), y=100*(cases.raw/test.raw)/max(cases.raw/test.raw)), col = "grey", size=1)+
  geom_point(data=data[!is.na(data$detected.roll),], aes(x=as.Date(dates2), y=100*detected.roll/max(detected.roll)), col = cols[5], size=1)+
  geom_point(data=data[!is.na(data$positivity.roll),],aes(x=as.Date(dates2), y=100*positivity.roll/max(positivity.roll)),size=1,col = "grey")+
  geom_line(data=data[!is.na(data$all),],aes(x=as.Date(dates2), y=100*all/max(all)),lwd=2, col = cols[5])+
  xlab("")+
  ylab("% of maximum")+
  ggtitle("Newfoundland and Labrador")+
  annotate("text", x = as.Date("2022-02-26"), y = 95, label = "positivity", col="grey")+
  annotate("text", x = as.Date("2022-02-20"), y = 45, label = "reported+unreported cases",fontface=2, col=cols[5])+
  annotate("text", x = as.Date("2022-02-24"), y = 32, label = "reported cases", col=cols[5])+
  theme_classic()+
  scale_x_date(date_breaks = "7 day", date_labels = "%d %b")+
  theme(axis.text.x = element_text(angle = 90),legend.position = "none")

gNL2=ggplot() +
 geom_point(data=tail(data,14), aes(x=as.Date(dates2), y=cases.raw), col = cols[5])+
  geom_line(data=tail(data,14), aes(x=as.Date(dates2), y=cases.raw), col = cols[5])+
  xlab("")+
  ylab("Daily average reported cases")+
  ggtitle("Newfoundland and Labrador (last 14 days)")+
  theme_classic()+
  ylim(c(0,300))+
  scale_x_date(date_breaks = "1 day", date_labels = "%d %b")+
  theme(axis.text.x = element_text(angle = 90),legend.position = "none")

gNL
gE+gC+gW+gL
```


<div class="alert alert-info">
In February 2022, 13$\%$ of tests reported by the Newfoundland and Labrador provincial PCR testing system have been positive, and on some days positivity has exceeded 20$\%$. _Could this high percentage of positive test results suggest that COVID-19 is increasing undetected in the Newfoundland and Labrador community?_

[Chui and Ndeffo-Mbah (2021)](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1009374) found that the prevalence of unreported cases during the previous week is approximated as the geometric mean of the per capita reported cases, and the proportion of tests that are positive, where these data are appropriately smoothed. I use this approximation to estimate the number of unreported cases a week ago, and prior. I graph the results to determine if the trend in **reported cases** (increasing, decreasing, or level) is consistent with the trend in **total cases (both reported and unreported)**.
</div>

```{r,echo=F}
gNL
```


**Figure 1.** Each of % positivity, total cases (reported and unreported), and reported cases are plotted relative to their maximum values, so that _trends_ can be compared. Recently, the percentage of tests that are positive, as reported by the provincial testing system, has been near its maximum value (grey dots; 14-day average). In contrast, reported cases have been declining or level since early January (blue dots; 14-day average as a % of the maximum value). Even with % positivity having recently been close to its maximum value, total COVID-19 cases (both reported and unreported) are relatively level (blue solid line; as a percentage of the maximum value). Reported cases and test positivity are used to calculate undetected cases 7 days ago, so total cases (blue solid line) can only be estimated up until 7 days before the last reported data. All data are from the [Newfoundland and Labrador COVID-19 Pandemic Update Hub](https://geohub-gnl.hub.arcgis.com/search?q=COVID).   

```{r,echo=F}
gE+gC+gW+gL
```

**Figure 2.** Results are similar at the Regional Health Authority level. There have been some times when total COVID-19 cases (both reported and unreported, solid lines) have increased for some Regional Health Authorities when reported cases (coloured dots) have not, but overall trends in reported and total (both reported and unreported; solid line) cases are similar. Methods and figure legend are as for Figure 1.


```{r,echo=F}
gNL2
```

**Figure 3.** Reported cases used for this analysis (last 14 days only). To account for days when cases and tests are not reported, the daily average is taken. The data are from the [Newfoundland and Labrador COVID-19 Pandemic Update Hub](https://geohub-gnl.hub.arcgis.com/search?q=COVID). 
